TEMPERATURES AND METALLICITIES FOR 10,000+ M DWARFS IN THE \apogee\ SURVEY

### 2018-08-10 paper/draft.tex
add some outline points

\newcommand{\apogee}{\textsl{APOGEE}}
\newcommand{\thecannon}{\textsl{The Cannon}}
\newcommand{\cannon}{\textsl{Cannon}}
\newcommand{\aspcap}{\textsl{ASPCAP}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\zmass}{\textsl{2MASS}}
\newcommand{\apass}{\textsl{APASS}}
\newcommand{\sdssiv}{\textsl{SDSS-IV}}

@@

\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; $\lambda/\Delta\lambda\sim22,500$, $1.5-1.7 \mathrm{\mu m}$) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H]) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim3550$K \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ dex we train a two-parameter model precise to 68K and 0.07 dex respectively. We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that \thecannon\ can infer spectral types to a precision of 0.6 types. Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey, and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and less biased than \aspcap\, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.

@@

% \begin{enumerate}
Very low mass (VLM) stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}.

Define what parameters/classifications are important for M dwarfs. We want to estimate empirical parameters/classifications (SPT, etc.), physical parameters (such as Teff, logg, metallicity), which we will refer to as \emph{stellar labels}.

@@

Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements.

The launch of the (\apogee\; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy. \color{gcolor}{What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't?}\color{black}

A number of M dwarf studies have already been conducted in \apogee\ (\citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs; \citealt{Souto:2017} -- modeled 2 M dwarfs, determined Teff/logg/metallicity + 13 elemental abundances; \citealt{Desphande:2013} -- modeled rotational and radial velocites for several hundred M dwarfs; \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity). \citealt{Gilhool:2018}; \citealt{Skinner:2018}.

@@

Why \thecannon\ approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016}) \color{gcolor}{For Hogg to articulate...}\color{black}

\begin{enumerate}
\item Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;

@@

\subsection{Test Sample \label{subsec:test_selection}} 
Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a sample of ``safe'' single, main-sequence M stars, with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):

@@

In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of \aspcap\ temperature values using a combination of \zmass\ and visual band photometries. \color{gcolor}{Brief description of the color-teff relations from \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}...}\color{black}. To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (\apass; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by \apass, and $JHK$ photometries from \apogee. In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the \aspcap\ temperatures for the same sources. \color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}

The gradients in the color magnitude diagrams of Figure \ref{fig:safe_selection} demonstrate the expected negative correlation between empirical \gaia\ $BP-RP$ colors and West-trained spectral classifications with the physical Mann-trained temperatures. Spectral type decreases nonlinearly with temperature... negatively correlated with metallcity for spectral types hotter than M3

@@

\begin{enumerate}
\item[-] Brief summary of what we did and found...

@@

\begin{figure}[ht]
\plotone{figures/spt_vs_teff_vs_feh.png}
\caption{West-trained spectral types vs. Mann-trained metallcities and temperatures. For each integer spectral type bin the mean temperatures and metallicities are shown as red error boxes. \label{fig:west_vs_mann_labels}}
\end{figure}

>>>>

\newcommand{\apogee}{\textsl{APOGEE}}
\newcommand{\thecannon}{\textsl{The Cannon}}
\newcommand{\cannon}{\textsl{Cannon}}
\newcommand{\aspcap}{\textsl{ASPCAP}}
\newcommand{\gaia}{\textsl{Gaia}}
\newcommand{\zmass}{\textsl{2MASS}}
\newcommand{\sdssiv}{\textsl{SDSS-IV}}

@@

\thecannon\ \citep{Ness:2015} is a flexible, data-driven spectral modeling and parameter inference framework, demonstrated on high-resolution Apache Point Galactic Evolution Experiment (\apogee; near-infrared) spectra of giant stars to estimate stellar labels (T$_{\rm eff}$, logg, [Fe/H]) to precisions higher than the model-grid pipeline. The lack of reliable stellar parameters reported by the \apogee\ pipeline for temperatures less than $\sim3550K$ \citep{Schmidt:2016}, motivates the extension of this approach to M dwarf stars. Using a training sample of 87 M dwarfs ranging $2859 < T_{\rm eff} < 4131$K, and $-0.48 <$ [Fe/H] $< 0.49$ we train a two-parameter model precise to 68K and 0.07 dex respectively. We then train a one-dimensional spectral type model using 51 M dwarfs ranging M0-M9 obtained from SDSS optical spectra, and demonstrate that \thecannon\ can infer spectral types to a precision of 0.6 types. Finally we apply our models to determine spectral subtypes, temperatures and metallicities for 10,311 M dwarfs in the survey and compare our results to \apogee\ pipeline measurements, and find our results to be significantly more precise and accurate than \aspcap, with scatters of $\sim80-100$K comapred to color-temperature relations from several literature sources.

@@

% \begin{enumerate}
Very low mass stars are the most abundant type of star, comprising $\sim 70 \%$ of the galaxy's population by number \citep{Bochanski:2010} and are known to have lifespans longer than the age of the universe (billions or up to trillions of years) \citep{Laughlin:1997} making them an interesting probe of local Milky way structure \citep{Bochanski:2007}. Relevance to exoplanet studies, and other reasons for scientific interest...

Advances in instrumentation and the lauch of several large scale spectroscopic stellar surveys launched in the past decade have dramatically increased the sample of known M dwarfs, including the \citealt{West:2011} catalog of 70,0841 sources from the Sloan Digital Sky Survey (SDSS DR7; \citealt{Abazajian:2009}), and the \citealt{Yi:2014} and \citealt{Guo:2015} catalogs of 93,619 from the Large Sky Area Multi-Object Fiber Spectroscopic Telescope (LAMOST DR1; \citealt{Zhao:2012}), which each have determined spectral subtype classifications, H$\alpha$ equivalent width, and a number of molecular band index measurements. Citations to any other large M dwarf catalogs that have

The launch of the (\apogee; \citealt{Majewski:2015}) survey as part of the Sloan Digital Sky Survey III mission \citealt{Eisenstein:2011} has introduced the largest sample of M dwarfs observed with high resolution spectroscopy. \color{gcolor}{What about \apogee\'s wavelength coverage, resolution, selection and other specs allow us to do things with M dwarfs that other surveys/instruments can't? What }\color{black}

Define what parameters/classifications are important for M dwarfs.

@@

A number of M dwarf studies have already been conducted in \apogee\, particularly towards measuring reliable fundamental atmospheric parameters and kinematic measurements:
\begin{enumerate}
\item \citealt{Schmidt:2016} -- analyzed trends and reliablility of ASPCAP parameters for K/M dwarfs
\item \citealt{Souto:2017} and \citealt{Souto:2018} -- modeled three exoplanet hosting M dwarfs (Kepler-138, Kepler-168 and Ross-128), determining Teff/logg/metallicity + 13 elemental abundances
\item \citealt{Desphande:2013} -- modeled rotational and radial velocities for 200+ M dwarfs, later extended by \citealt{Gilhool:2018} to 700+ M dwarfs
\item \citealt{Rajpurohit:2018} -- tested BTSettl model grids on 45 M dwarfs, estimated Teff/logg/metallicity)
\item \citealt{Skinner:2018} - identified and measured mass ratios and radial velocities for 44 M dwarf binaries
\end{enumerate}

@@

Why \thecannon\ approach? brief importance statment cannon and what it does (cite \citealt{Ness:2015}, \citealt{Ho:2017a}, \citealt{Casey:2016})

\begin{enumerate}
\item This paper will refer to any physical or empirical measurement of a star as \emph{stellar label}

\item Can be used to amplify a small number of labeled M-dwarfs in a training set into a huge number of labeled M-dwarfs, provided that there are overlapping objects;

@@

\subsection{Test Sample \label{subsec:test_selection}} 
Out of the total \apogee\ DR14 catalog of 258,475 sources, we selected 254,478 sources which were in the cross match of \textsl{Gaia DR2} (\citealt{Brown:2018}) and applied \gaia\ color-magnitude cuts of $1<BP-RP<6$ and $7.5<M_G<20$, yielding a sample of 14,827 sources. From there we applied the following selection procedure to identify a sample of single, main-sequence M stars. Although our catalog reports the temperatures, metallicities, and spectral types for XXX sources out of 14,827 with $\chi^2<200,000$, the sources we deem ``safe'', with minimal contamination from reddened K dwarfs, pre-main sequence stars, and binaries, are selected by the following criteria (see Figure \ref{fig:cmd_selection}):

@@

In \citealt{Schmidt:2016} the authors used three different color-temperature relations to check the consistency of \aspcap\ temperature values using a combination of \zmass\ and visual band photometries. \color{gcolor}{Brief description of the color-teff relations from \citealt{Casagrande:2008}, \citealt{Boyajian:2013}, and \citealt{Mann:2015}...}\color{black}.

To obtain visual band magnitudes for a set of sources, we cross-matched the 10,311 sources in our ``safe'' test sample to the AAVSO Photometric All-Sky Survey DR9 (APASS; \citealt{Henden:2016}), to obtain a subsample of 6,000 sources with both $BV$ photometries measured by APASS, and $JHK$ photometries from \apogee. In Figure \ref{fig:aspcap_cannon_validation} we repeat the temperature validation tests for each of the three relations and compare the \aspcap\ temperatures for the same sources. \color{gcolor}{Comments about validation results and the uncertainty of our Cannon temperatures...}\color{black}

@@

\begin{enumerate}
\item[-] Brief summary of what we did and found

@@

\begin{figure}[ht]
\plotone{figures/spt_vs_teff_vs_feh.png}
\caption{West-trained spectral types vs. Mann-trained metallcities and temperatures. For each integer spectral type bin the mean temperatures and metallicities are shown as red error boxes. \label{fig:west_vs_mann}}
\end{figure}
